Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
paper full

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$.\MANUEL{necesitamos las ecuaciones? Swap de una permutación es simplemente la distancia de kendall de la permutación inversa. .}
\EKHINE{asi mas claro, no se van a perder con esto}
 %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

@@

%During our experiments, we realized that the \emph{swap} distance used by CEGO
%assumes rankings, i.e., it actually measures the \ken distance, yet CEGO did
% not invert the permutation before evaluating the PFSP.
During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP.\MANUEL{no me convence esta forma de expresarlo porque la crítica obvia sería que por qué invertimos la función objectivo y no la función de distancia?} \EKHINE{es lo mismo}We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. Of course, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. Therefore, in the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

@@

\myparagraph{Comparison of the best solutions found}
%
We now turn our analysis to the best (minimum) fitness found at the end of the
run by the two best algorithms, that is, \CEGOinv and UMM$_{400}$. We show in
Table~\ref{tab:results} the mean fitness (and standard deviation) of the best
solution found over the 10 runs of each algorithm, together with a 95\%
confidence interval (CI) around their mean fitness difference. Except for PFSP
instances \texttt{rec05} and \texttt{rec31}, the differences are statistically
significant in favour of \CEGOinv (the intervals do not contain the value
zero). Nevertheless, the CIs indicate that the differences, although
consistent, are relatively small.

@@

Table~\ref{tab:results} shows a brief comparison of time and the full ones are given in Table~\ref{tab:time}. The difference is time in impressive. 
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
For CEGO in both versions the average time for 400 evaluations is more than 22 hour. This is 300 times slower than UMM for the same number of function evaluations. 
Moreover, UMM scales good for larger instances in quality performance. 
The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \ekhine{MANUEL, tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-04} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

>>>>

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$. %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

@@

%During our experiments, we realized that the \emph{swap} distance used by CEGO
%assumes rankings, i.e., it actually measures the \ken distance, yet CEGO did
% not invert the permutation before evaluating the PFSP.
During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP.
%\MANUEL{no me convence esta forma de expresarlo porque la crítica obvia sería que por qué invertimos la función objectivo y no la función de distancia?} \EKHINE{es lo mismo}
We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. 
Note that using the Swap distance as claimed by CEGO is equivalent to using \ken distance and inverting the permutation before calling the objective
function. Thus, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. In the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

@@

\myparagraph{Comparison of the best solutions found}
%
We now turn our analysis to the best (minimum) fitness found at the end of the
run by the two best algorithms, that is, \CEGOinv and UMM$_{400}$. We show in
Table~\ref{tab:results} the mean fitness (and standard deviation) of the best
solution found over the 10 runs of each algorithm, together with a 95\%
confidence interval (CI) around their mean fitness difference. 
The 
Except for PFSP
instances \texttt{rec05} and \texttt{rec31}, the differences are statistically
significant in favour of \CEGOinv (the intervals do not contain the value
zero). Nevertheless, the CIs indicate that the differences, although
consistent, are relatively small.

@@

Table~\ref{tab:results} shows a brief comparison of time and the full ones are given in Table~\ref{tab:time}. The difference is time in impressive. 
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
For CEGO in both versions the average time for 400 evaluations is more than 22 hour. This is 300 times slower than UMM for the same number of function evaluations. 
Moreover, UMM scales good for larger instances in quality performance. 
The computational complexity of UMM is $O(n \log n)$ at each iteration which causes the abrupt differencein the time performance of the algorithms. The details can be seen in Table~\ref{tab:results_time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-02} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59d11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59d11xx}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}
